High-resolution ALMA observations of V4046\,Sgr: a circumbinary disc with a thin ring

### 2021-06-22 Draft_V4046Sgr_mnras.tex
Updates from Overleaf

% Abstract of the paper
\begin{abstract}
    The nearby V4046\,Sgr spectroscopic binary hosts a gas-rich disc known for its wide cavity and dusty ring. We present high resolution ($\sim$20\,mas or 1.4\,au) ALMA observations of the 1.3\,mm continuum of V4046\,Sgr which, combined with SPHERE--IRDIS polarized images and well-sampled spectral energy distribution (SED), allow us to propose a physical model using radiative transfer (RT) predictions. The new ALMA data reveal a thin ring at a radius of 13.15$\pm$0.42\,au (Ring13), with a radial width of 2.46$\pm$0.56\,au. Ring13 is surrounded by a $\sim$10\,au-wide gap, and it is flanked by a mm-bright outer ring (Ring24) with a sharp inner edge at 24\,au. Between 25 and $\sim$35\,au the brightness of Ring24 is relatively flat and then breaks into a steeper tail that reaches out to $\sim$60\,au. In addition, central emission is detected close to the star which we interpret as a tight ring made of dust grains with a lower size limit of 0.5\,mm. In order to reproduce the SED, the model also requires an inner ring at $\sim$5\,au (Ring5) composed mainly of small dust grains, hiding under the IRDIS coronagraph, and surrounding an inner circumbinary disc. The surprisingly thin Ring13 is nonetheless roughly 10 times wider than its expected vertical extent. The strong near-far disc asymmetry at 1.65\,$\micron$ points at a very forward-scattering phase function, and requires grain radii of no less than 0.4\,$\micron$. The previously reported scattered light shadow of the secondary star is also reproduced by our model.
\end{abstract}

@@

Building  a parametric model that simultaneously fits the near-IR scattered light and millimetre continuum emission  can provide a way to link the observations to the structures in the underlying dust and gas density distributions, and shed light on the complex processes that shaped them. In this paper, we present new ALMA observations of the continuum emission at 1.3\,mm of V4046\,Sgr, which reveal a new inner ring in the disc. We also reproduce these observations with a 3D parametric model that fits all available data, including the polarized scattered light SPHERE image, the spectral energy distribution (SED), as well as the new high definition 1.3\,mm continuum map. Section~\ref{sec:Observations} describes the available observations that we aim to model. Section~\ref{sec:model} describes the structural parameters of our model. A description of our parameter space exploration can be found in Appendix~\ref{sec:Appendix}. The results of our modelling are presented in Section~\ref{sec:results}, including a discussion of the main findings. Finally, in Section~\ref{sec:Conclusions} we present our conclusions.

@@

The resulting {\tt uvmem} image shown in the top right panel of Fig.~\ref{fig:images_vs_simulated} exhibits previously unseen substructure of the disc. This image reveals that the disc features two rings of large dust grains with a broad gap between them, i.e. Ring13 at 13\,au and Ring24 starting at 24\,au. The wide and bright Ring24 reaches its peak intensity at $\sim$30\,au, continues relatively flat and then breaks at $\sim$35\,au into a steeper tail. While this is the first observation of Ring13, \citet{Ru_z_Rodr_guez_2019} did anticipate its existence as their ALMA continuum image showed a distinct excess between $\sim$10 and 17\,au.

Ring13 is surprisingly narrow and seems to be off-centred relative to the Gaia stellar position, at the origin of coordinates in Fig.~\ref{fig:images_vs_simulated}. We determined the ring's centre and orientation by minimising the dispersion of the disc radial profiles from 6 to 18\,au, by using the {\tt MPolarMaps} package described in Casassus et al. (2021, in prep). The resulting disc orientation is set at a position angle (PA) of 77.31$\pm$0.03\,$\degr$ (east of north), with an inclination of 32.96$\pm$0.02\,$\degr$, and the optimal ring centre is at $\Delta \alpha = -4\pm0.02$\,mas $\Delta \delta = 13\pm0.05$\,mas relative to the Gaia position of the stars. The offset of the center of the cavity and the nominal stellar positions is entirely consistent with a pointing error, since the pointing accuracy of ALMA is $\sim$5\,mas for the signal to noise ratio of our image (calculated using the ALMA Technical Handbook).

In a second optimization of the disc orientation, but this time aiming for Ring24 with a radial domain from 20\,au to 70\,au, we obtained a PA of 76.86$\pm$0.02\,$\degr$, with an inclination of 33.92$\pm$0.01\,$\degr$ and a centre at $\Delta \alpha = 2\pm0.02$\,mas $\Delta \delta = 12\pm0.02$\,mas relative to the stars. We see that both Ring13 and Ring24 share a very similar orientation and centre, given the errors and the pointing error. However, quite likely due to the joint effect of all these small differences, we can see that there are some hints for a somewhat different orientation. This is summarised in the second panel of Fig.~\ref{fig:polarring}, where it shows the centroids of the Gaussian fitting at each azimuth for both optimization. The curve that follows these centroids fitted for each of the optimizations goes almost entirely within the errors of the other one, leaving the window open for discussion as to whether these differences are just a consequence of the pointing error or are a true feature of the disc. Future imaging and further research will clarify the doubts.

The width of Ring13 can be measured in polar coordinates by fitting 1-D Gaussians at each azimuth, thus obtaining a width and centroid. The top panel of Fig.~\ref{fig:polarring} shows the positions of the obtained centroids and widths over the polar decomposition of the ALMA image. On average, we obtained a FWHM of 2.83$\pm$0.50\,au, and a stellocentric radius of 13.15$\pm$0.42\,au. As the {\tt uvmem} model image has an approximate {\tt uvmem} beam of $\sim$0$\farcs021 \,\times \,\sim$0$\farcs018$ (or $\sim$1.4\,au at 71.1\,pc), we see that Ring13 is resolved. After subtraction of the {\tt uvmem} beam, the ring width is $\sim$2.46$\pm$0.56\,au.

Interestingly, the ALMA image also detects 1.3\,mm continuum emission near the stellar positions (see the inset in Fig.~\ref{fig:images_vs_simulated}). Since this central emission is larger than the angular resolution, it is probably stemming from thermal emission from large dust grains rather than directly from the stars. The peak intensity of this dust structure is at an estimated distance of only $0\farcs012\pm0\farcs002$, or $\sim$0.85$\pm$0.14\,au from the binary system, and, estimating its form as a Gaussian ellipse, it has a mean FWHM of $\sim$1.43\,au. Measurements of this central emission made by \citet{Francis_2020}, assume a power-law profile for an inner disc that goes from 0.1 to 4\,au, and has a total dust mass of 0.013$\pm$0.002\,M$_{\earth}$.

@@

\begin{figure*}
  \includegraphics[width=\textwidth]{hot_two_E.pdf}
  \caption{Comparison of observations (top) and simulated images (bottom) at 1.65\,$\micron$ (left) and 1.3\,mm continuum (right) of the circumbinary disc orbiting V4046\,Sgr. \textit{Top left panel:} SPHERE--IRDIS \textit{H}-band image with a yellow filled circle that represents the N\_ALC\_YJH\_S coronagraph with a radius of $\sim$0$\farcs12$, or $\sim$8.6\,au at 71.48\,pc. \textit{Top right panel:} 1.3\,mm continuum {\tt uvmem} model image. The small orange ellipse shows an estimated {\tt uvmem} beam size ($\sim$0$\farcs021 \, \times \, \sim$0$\farcs018$). The inset zooms into the central emission, and the green star marks the center of the inner ring. \textit{Bottom left panel:} synthetic image at 1.65\,$\micron$. \textit{Bottom right panel:} synthetic image at 1.3\,mm convolved with the {\tt uvmem} beam. For all the images in the figure the colour scale is linear, and the color bar on the right applies only to the right images.}
  \label{fig:images_vs_simulated}
\end{figure*}

@@

The binary phase reported by \citet{dOrazi} in the scattered light observation is at a PA of 265\,$\degr$, east of north. Using this measurement, the binary phase was calculated at the time of the ALMA observation at a PA of $\sim$80\,$\degr$. There is no hint of radio decrements in either Ring13 nor in Ring24 that would match shadowing at this PA. This could be explained by inefficient disc cooling compared to the speed of the illumination pattern \citep[see ][ for estimates of this cooling timescale]{Casassus2019MNRAS.486L..58C}.

@@

The observed SED was collected from data in the literature \citep{1988iras....7.....H, 1990A&A...234..230H, Jensen_97, 2000A&A...355L..27H, 2001KFNT...17..409K, 2003yCat.2246....0C, 2007PASJ...59S.369M, 2008PASP..120.1128O, 2010A&A...514A...1I, 2012yCat.2311....0C}, available online in \textsc{vizier}. We also used archival \textit{Spitzer} IRS spectroscopic data available in the CASSIS database \citep{Lebouteiller_2015} (see Fig.~\ref{fig:SED}). The SED describe the characteristic “dip” near 10 $\micron$, a common element of transition disc, as described by \citet{Rosenfeld_2013}. \citet{Jensen_97} concluded that these data matched that of a extended circumbinary disc truncated at $\sim$0.2\,au, as the interior would also be expected to be cleared by dynamical effects of the central binary \citep{Art_Lu}.

The SED reveals a small near--infrared (NIR) excess of 0.9$\pm$3.7\,\% \citep{Francis_2020}. The mm-bright central emission, on the other hand, suggests a much more massive inner dust disc. The lack of correlation between the excess of NIR fluxes and mm emission could point to a lack of micron-sized dust in this central ring. This could be due to efficient dust growth or, as proposed by \citet{Francis_2020}, due to an incongruity between the sublimation radius and the inner boundary of the disc.

@@

The multi-frequency data can be interpreted in terms of a physical structure using radiative transfer predictions, for which we used the \textsc{radmc3d} package \citep[version 2.0,][]{Dullemond_2012}. The general framework of the parametric model that we developed is similar to that in \citet{2018MNRAS.477.5104C} for DoAr\,44, and the initial model values were inspired from those in \citet{Rosenfeld_2013}, \citet{Ru_z_Rodr_guez_2019} and \citet{2019ApJ...882..160Q}. Computing a high-resolution radiative transfer model that succeeds in the reproduction of multi-frequency imaging and the SED, is finding a solution to a highly degenerative problem, so obtaining an ideal model with a full parameter exploration requires a level of computation and time that exceeds our capabilities and the scope of this paper. Consequently, our approach was to find through trial and error a set of values for the parameters that closely fit the available data, and then, improve this fit by implementing one dimensional chi squared optimizations for some key parameters (more on this in Sec.~\ref{sec:results}). The final structure of the parametric model is summarised in Fig.~\ref{fig:profiles}, where we show on the top panel the surface density profiles for gas and dust grains populations, and in the lower panel the respective aspect ratio profile.

The stars were modeled using two Kurucz photosphere models \citep{1979ApJS...40....1K, 1997A&A...318..841C}, with T$_{\mathrm{eff},1} =$ 4350\,K, R$_{*,1} =$ 1.064\,R$_{\sun}$, M$_{*,1} =$ 0.90\,M$_{\sun}$ and T$_{\mathrm{eff},2} =$ 4060\,K, R$_{*,2} =$ 1.033\,R$_{\sun}$, M$_{*,2} =$ 0.85\,M$_{\sun}$, respectively and with an accretion rate of $\mathrm{log}(\,\dot{\mathrm{M}}/(\mathrm{M}_{\sun}\,\mathrm{yr^{-1}})) = -$9.3 for both cases to include excess UV due to stellar accretion \citep{10.1111/j.1365-2966.2011.19366.x}. The stars were placed at a mutual separation of 0.041\,au, so that their centre of mass coincides with the origin. Their PA was set to 250\,$\degr$, so that the secondary lies to the NW from the primary, thus casting the same shadow as observed by \citet{dOrazi}.

Reproducing the radial and vertical structure of the V4046\,Sgr disc turned out to be challenging. We built the model in terms of the gas distribution, and with two main dust populations: larger grains that are vertically settled and dominate the total dust mass, and a population of smaller grains that are uniformly mixed with the gas and reach higher regions above the mid-plane. We also added a third dust population of larger grains with a tightly confined distribution and a special size range just to account for the central emission.

@@

The parametric scale height profiles for the gas and for each dust population are 
\begin{equation}
    \label{scale}
  H(r)=\chi \, H_{\circ} \,\left( \frac{r}{r_{\circ}}\right)^{1+\psi(r)},
\end{equation}
where $H_\circ$ is the scale height at $r$ = $r_\circ$, $\psi$ is the flaring index and $\chi$ is a scaling factor (in the range $0-1$) that mimics dust settling. In hydrostatical equilibrium, dust diffusion and settling are expected to balance each other, leading to a settling factor of $\chi = \sqrt{D_{\rm d}/(D_{\rm d} + St)}$ \citep{Dubrulle1995}. Here, $D_{\rm d}$ is a dimensionless parameter that informs about the level of diffusivity \citep[it is typically assumed to be similar to the level of turbulence for the particle sizes regarded here][]{2007Icar..192..588Y}. The Stokes number,
\begin{equation}
    {\rm St} := \frac{1}{2} \,\frac{\pi a \rho_{\rm mat}}{\Sigma_{\rm gas}}, 
\end{equation}
summarises the dynamical behaviour of a particle in a given environment (where $\rho_{\rm mat}$ represents a dust particle's material density). By definition, the gas has no settling, and the small dust grains settling is negligible, so that $\chi_{\rm gas}=\chi_{\rm sd}=1$. As we do not know the level of diffusivity in the disc (this will be a topic of interest in Section~\ref{sec:results}), nor the exact value of $\Sigma_{\rm gas}$, we infer the scaling factor for large dust, $\chi_{\rm ld}$, from the width of Ring13. In the radial profile this ring is observed two to three times wider in the gas-tracing NIR than in the fluxes received from larger grains by ALMA. We assume that the same ratio holds in the vertical direction (due to the settling of larger grains towards the mid-plane of the disc), leading to $\chi_{\mathrm{ld}}=0.4$. This is analogous to assuming equal radial and vertical turbulent diffusions.

For the vertical structure, \citet{dOrazi} found flaring angles (i.e. height of the ring over the disc midplane divided by the radius of the ring), of $\varphi$ = 6.2$\pm$0.6\,$\degr$ for the inner ring and $\varphi$ = 8.5$\pm$1.0\,$\degr$ for the outer one. Our model aims to reproduce those values by using two different flaring indices, $\psi_1$ and $\psi_2$. The separation between the two values was set at $r = 18$\,au with $\psi_1=0.2$ for Ring5 and Ring13, and $\psi_2=0.5$ for Ring24. The scale height is set to $H_\circ = 0.89$\,au at $r_\circ = 18$\,au. This vertical structure is consistent with the measurements made by \citet{dOrazi}, as the model flaring angles are $\varphi_{\mathrm{inner}}$ = 6.4\,$\degr$ and $\varphi_{\mathrm{outer}}$ = 7.6\,$\degr$. This is summarized in bottom panel of Fig.~\ref{fig:profiles} that shows the aspect ratio profile $h(r)=H(r)/r$.

Although both ALMA and SPHERE--IRDIS images display two-ringed morphologies, we propose a three-ringed structure plus an inner disc to reproduce the observations. This bold decision is due the major fit improvement in the SED and the polarize image (more on this in Sec.~\ref{sec:results}). We separate the disc into four individual regions: the inner disc, Ring5, Ring13, and Ring24. The combined gas surface density profile is then given by:
\begin{equation}
  \Sigma_{\mathrm{gas}}(r) = \Sigma_{\mathrm{inner\,disc}} + \Sigma_{\mathrm{Ring5}} + \Sigma_{\mathrm{Ring13}} + \Sigma_{\mathrm{Ring24}}
\end{equation}

@@

We chose to use Gaussian profiles to parametrize Ring5 and Ring13,
\begin{equation}
  \Sigma_{\mathrm{ring}}(r) = \frac{\Sigma_\circ}{\sqrt{2 \pi} \sigma}
  \, \mathrm{exp}\left[-\frac{1}{2}\left(\frac{r-\mu}{\sigma}\right)^{2}\right],
\end{equation}
where we define constants that correspond to the centroid radii \{$\mu_{\mathrm{Ring5}}=5.2$\,au, $\mu_{\mathrm{Ring13}}=14.9$\,au\}, ring widths \{$\sigma_{\mathrm{Ring5}}=0.25$\,au, $\sigma_{\mathrm{Ring13}}=2.26$\,au\}, and normalizations \{$\Sigma_{\circ,\mathrm{Ring5}}=3.3$\,$\mathrm{g\,cm^{-3}}$, $\Sigma_{\circ,\mathrm{Ring13}}=6.0 \times 10^{-1}$\,$\mathrm{g\,cm^{-3}}$\} for both components separately.

For Ring24 we used the same power-law as for the inner disc but scaled by an empirically obtained factor, $\delta_{\mathrm{sd}}(r)$ and by $\epsilon(r)$, a parameter that allows us model a smoother inner edge of the outer ring:
\begin{equation}
  \Sigma_{\mathrm{Ring24}}(r) = \Sigma_{\mathrm{inner\,disc}}(r)\, \delta_{\mathrm{sd}}(r)\,\epsilon(r),
\end{equation}
with $\delta_{\mathrm{sd}}(r) = 1.0\times 10^5$ for $r > 18$\,au and zero for lower radii, and
%\begin{equation}
%  \delta_{\mathrm{sd}}(r) =
%  \begin{cases}
%  0    & r < 18\text{\,au} \\
%  4.2\times 10^4 & r > 18\text{\,au},
%  \end{cases}
%\end{equation}
\begin{equation}
    \epsilon(r) = 
    \begin{cases}
  1    & r < R_\mathrm{in} \,\, {\rm and} \,\, R_\mathrm{peak} < r \\
  \left(\frac{ r - R_\mathrm{in}}{R_\mathrm{peak} - R_\mathrm{in}}\right)^3 & R_\mathrm{in} < r < R_\mathrm{peak},
    \end{cases}
\end{equation}
where $R_\mathrm{in}$ and $R_\mathrm{peak}$ respectively mark the inner edge and the location of maximum density of the outer ring. We used $R_\mathrm{in,gas}$ = 18.0\,au and $R_\mathrm{peak,gas}$ = 26.4\,au.

The total dust-to-gas mass ratio is taken to be $\zeta = 0.047$ \citep[as in][]{Rosenfeld_2013}. The small dust grains are assumed to only make up for a fraction of $f_\mathrm{sd}=1\%$ of the total dust mass. As small dust is typically tightly coupled to the gas dynamics, its density profile is expected to follow the gas density. Then the density of small dust can be calculated as:
\begin{equation}
\rho_{\mathrm{small-dust}}(r,z)=\rho_{\mathrm{gas}}(r,z)\, f_{\mathrm{sd}} \: \zeta .
\end{equation}

>>>>

% Abstract of the paper
\begin{abstract}
    The nearby V4046\,Sgr spectroscopic binary hosts a gas-rich disc known for its wide cavity and dusty ring. We present high resolution ($\sim$20\,mas or 1.4\,au) ALMA observations of the 1.3\,mm continuum of V4046\,Sgr which, combined with SPHERE--IRDIS polarized images and a well-sampled spectral energy distribution (SED), allow us to propose a physical model using radiative transfer (RT) predictions. The new ALMA data reveal a thin ring at a radius of 13.15$\pm$0.42\,au (Ring13), with a radial width of 2.46$\pm$0.56\,au. Ring13 is surrounded by a $\sim$10\,au-wide gap, and it is flanked by a mm-bright outer ring (Ring24) with a sharp inner edge at 24\,au. Between 25 and $\sim$35\,au the brightness of Ring24 is relatively flat and then breaks into a steeper tail that reaches out to $\sim$60\,au. In addition, central emission is detected close to the star which we interpret as a tight circumbinary ring made of dust grains with a lower size limit of 0.8\,mm at 1.1\,au. In order to reproduce the SED, the model also requires an inner ring at $\sim$5\,au (Ring5) composed mainly of small dust grains, hiding under the IRDIS coronagraph, and surrounding the inner circumbinary disc. The surprisingly thin Ring13 is nonetheless roughly 10 times wider than its expected vertical extent. The strong near-far disc asymmetry at 1.65\,$\micron$ points at a very forward-scattering phase function, and requires grain radii of no less than 0.4\,$\micron$. The previously reported scattered light shadow of the secondary star is also reproduced by our model.
\end{abstract}

@@

Building a parametric model that simultaneously fits the NIR scattered light and millimetre continuum emission  can provide a way to link the observations to the structures in the underlying dust and gas density distributions, and shed light on the complex processes that shaped them. In this paper, we present new ALMA observations of the continuum emission at 1.3\,mm of V4046\,Sgr, which reveal a new inner ring in the disc. We also reproduce these observations with a 3D parametric model that fits all available data, including the polarized scattered light SPHERE image, the spectral energy distribution (SED), as well as the new high definition 1.3\,mm continuum map. Section~\ref{sec:Observations} describes the available observations that we aim to model. Section~\ref{sec:model} describes the structural parameters of our model. A description of our parameter space exploration can be found in Appendix~\ref{sec:Appendix}. The results of our modelling are presented in Section~\ref{sec:results}, including a discussion of the main findings. Finally, in Section~\ref{sec:Conclusions} we present our conclusions.

@@

The resulting {\tt uvmem} image shown in the top right panel of Fig.~\ref{fig:images_vs_simulated} reveals new substructure of the disc, in the form of two rings of large dust grains with a broad gap between them, i.e. Ring13 at 13\,au and Ring24 starting at 24\,au. The wide and bright Ring24 reaches its peak intensity at $\sim$30\,au, continues relatively flat and then breaks at $\sim$35\,au into a steeper tail. While this is the first observation of Ring13, \citet{Ru_z_Rodr_guez_2019} anticipated its existence as their ALMA continuum image showed a distinct excess between $\sim$10 and 17\,au.

Ring13 is surprisingly narrow and seems to be off-centred relative to the Gaia stellar position, at the origin of coordinates in Fig.~\ref{fig:images_vs_simulated}. We determined the ring's centre and orientation using the {\tt MPolarMaps} package described in \citep{2021arXiv210408379C}, which minimises the dispersion of the intensity radial profiles from 6 to 18\,au. The resulting disc orientation is set at a position angle (PA) of 77.31$\pm$0.03\,deg (east of north), with an inclination of 32.96$\pm$0.02\,deg, and the optimal ring centre is at $\Delta \alpha = -4\pm0.02$\,mas $\Delta \delta = 13\pm0.05$\,mas relative to the Gaia position of the stars. The offset of the center of the cavity and the nominal stellar positions are coincident within the pointing accuracy of ALMA, which  is $\sim$5\,mas for the signal to noise ratio of our image (calculated using the ALMA Technical Handbook).

In a second optimization of the disc orientation, but this time aiming for Ring24 with a radial domain from 20\,au to 70\,au, we obtained a PA of 76.86$\pm$0.02\,deg, with an inclination of 33.92$\pm$0.01\,deg and a centre at $\Delta \alpha = 2\pm0.02$\,mas $\Delta \delta = 12\pm0.02$\,mas relative to the stars. We see that both Ring13 and Ring24 share a very similar orientation and centre, given the errors and the pointing accuracy. However, see that there are some hints for a somewhat different orientation in the azimuthal profiles for the ring radii, which may nonetheless be accounted for by the joint effect of all these small differences in disc orientation. This is summarised in Fig.~\ref{fig:polarring}b, which characterises the radial position and width of Ring13 as a function of azimuth by using radial Gaussian fits at each azimuth. The radial centroids  for both disc orientation overlap within the errors, but there is a systematic trend for the difference between the two. It may be that this small difference reflects a finite  intrinsic eccentricity of one or both rings. Deeper imaging is required to progress on this question.

On average, we obtained a radial FWHM for Ring13 of 2.83$\pm$0.50\,au, and a stellocentric radius of 13.15$\pm$0.42\,au. As the {\tt uvmem} model image has an approximate {\tt uvmem} beam of $\sim$0$\farcs021 \,\times \,\sim$0$\farcs018$ (or $\sim$1.4\,au at 71.1\,pc), we see that Ring13 is resolved. After subtraction of the {\tt uvmem} beam, the ring width is $\sim$2.46$\pm$0.56\,au.

Interestingly, the ALMA image also detects 1.3\,mm continuum emission near the stellar positions (see the inset in Fig.~\ref{fig:images_vs_simulated}). Since this central emission is larger than the angular resolution, it is probably stemming from thermal emission from large dust grains rather than directly from the stars. The peak intensity of this dust structure is at an estimated distance of only $0\farcs012\pm0\farcs002$, or $\sim$0.85$\pm$0.14\,au from the binary system, and, estimating its form as a Gaussian ellipse, it has a mean FWHM of $\sim$1.43\,au.

@@

\begin{figure*}
  \includegraphics[width=\textwidth]{hot_two_E.pdf}
  \caption{Comparison of observations (top) and simulated images (bottom) at 1.65\,$\micron$ (left) and 1.3\,mm continuum (right) of the circumbinary disc orbiting V4046\,Sgr. \textit{Top left panel:} SPHERE--IRDIS \textit{H}-band image with a yellow filled circle that represents the N\_ALC\_YJH\_S coronagraph with a radius of $\sim$0$\farcs12$, or $\sim$8.6\,au at 71.48\,pc. \textit{Top right panel:} 1.3\,mm continuum {\tt uvmem} model image. The small orange ellipse shows an estimated {\tt uvmem} beam size ($\sim$0$\farcs021 \, \times \, \sim$0$\farcs018$). The inset zooms into the central emission, and the green star marks the center of the inner ring. \textit{Bottom left panel:} synthetic image at 1.65\,$\micron$. \textit{Bottom right panel:} synthetic image at 1.3\,mm convolved with the {\tt uvmem} beam. For all the images in the figure the colour scale is linear, and the color bar on the right applies only to the images on the right-hand panel.}
  \label{fig:images_vs_simulated}
\end{figure*}

@@

The binary phase reported by \citet{dOrazi} in the scattered light observation is at a PA of 265\,deg, east of north. Using this measurement, the binary phase was calculated at the time of the ALMA observation at a PA of $\sim$80\,deg.

@@

The observed SED was collected from data in the literature \citep{1988iras....7.....H, 1990A&A...234..230H, Jensen_97, 2000A&A...355L..27H, 2001KFNT...17..409K, 2003yCat.2246....0C, 2007PASJ...59S.369M, 2008PASP..120.1128O, 2010A&A...514A...1I, 2012yCat.2311....0C}, available online in \textsc{vizier}. We also used archival \textit{Spitzer} IRS spectroscopic data available in the CASSIS database \citep{Lebouteiller_2015} (see Fig.~\ref{fig:SED}). The SED exhibits the dip near 10\,$\micron$ characteristic of transition discs, as described by \citet{Rosenfeld_2013}. \citet{Jensen_97} concluded that these data matched that of a extended circumbinary disc truncated at $\sim$0.2\,au, as the interior would also be expected to be cleared by dynamical effects of the central binary \citep{Art_Lu}.

@@

The multi-frequency data can be interpreted in terms of a physical structure using radiative transfer predictions, for which we used the \textsc{radmc3d} package \citep[version 2.0,][]{Dullemond_2012}. The general framework of the parametric model that we developed is similar to that in \citet{2018MNRAS.477.5104C} for DoAr\,44, and the initial model values were inspired from those in \citet{Rosenfeld_2013}, \citet{Ru_z_Rodr_guez_2019} and \citet{2019ApJ...882..160Q}. A high-resolution radiative transfer model that reproduces multi-frequency imaging and the SED, is a solution to a highly degenerate problem, so a full parameter exploration requires a level of computation and time that exceeds our capabilities and the scope of this paper. Consequently, our approach was to find through trial and error a set of values for the parameters that closely fit the available data, and then, improve this fit by implementing one dimensional least squared optimizations for some key parameters (more on this in Sec.~\ref{sec:results}). The final structure of the parametric model is summarised in Fig.~\ref{fig:profiles}, where we show on the top panel the surface density profiles for gas and dust grains populations, and in the lower panel the respective aspect ratio profile.

The stars were modeled using two Kurucz photosphere models \citep{1979ApJS...40....1K, 1997A&A...318..841C}, with T$_{\mathrm{eff},1} =$ 4350\,K, R$_{*,1} =$ 1.064\,R$_{\sun}$, M$_{*,1} =$ 0.90\,M$_{\sun}$ and T$_{\mathrm{eff},2} =$ 4060\,K, R$_{*,2} =$ 1.033\,R$_{\sun}$, M$_{*,2} =$ 0.85\,M$_{\sun}$, respectively and with an accretion rate of $\mathrm{log}(\,\dot{\mathrm{M}}/(\mathrm{M}_{\sun}\,\mathrm{yr^{-1}})) = -$9.3 for both cases to include excess UV due to stellar accretion \citep{10.1111/j.1365-2966.2011.19366.x}. The stars were placed at a mutual separation of 0.041\,au, so that their centre of mass coincides with the origin. Their PA was set to 250\,deg, so that the secondary lies to the NW from the primary, thus casting the same shadow as observed by \citet{dOrazi}.

Reproducing the radial and vertical structure of the V4046\,Sgr disc turned out to be challenging. We built the model in terms of the gas distribution, and with two main dust populations: larger grains that are vertically settled and dominate the total dust mass, and a population of smaller grains that are uniformly mixed with the gas and reach higher regions above the mid-plane. We also added a third dust population of larger grains with a  distribution tightly confined to the stellar vicinity, within 2\,au, and a special size range meant to account for the central emission.

@@

The parametric scale height profiles for the gas and for each dust population are 
\begin{equation}
    \label{scale}
  H(r)=\chi \, H_{\circ} \,\left( \frac{r}{r_{\circ}}\right)^{1+\psi(r)},
\end{equation}
where $H_\circ$ is the scale height at $r$ = $r_\circ$, $\psi$ is the flaring index and $\chi$ is a scaling factor (in the range $0-1$) that mimics dust settling. In hydrostatical equilibrium, dust diffusion and settling are expected to balance each other, leading to a settling factor of $\chi = \sqrt{D_{\rm d}/(D_{\rm d} + St)}$ \citep{Dubrulle1995}. Here, $D_{\rm d}$ is a dimensionless parameter that informs about the level of diffusivity \citep[it is typically assumed to be similar to the level of turbulence for the particle sizes regarded here][]{2007Icar..192..588Y}. The Stokes number,
\begin{equation}
    {\rm St} = \frac{1}{2} \,\frac{\pi a \rho_{\rm mat}}{\Sigma_{\rm gas}}, 
\end{equation}
summarises the dynamical behaviour of a particle in a given environment (where $\rho_{\rm mat}$ represents a dust particle's material density). By definition, the gas has no settling, and the small dust grains settling is negligible, so that $\chi_{\rm gas}=\chi_{\rm sd}=1$. As we do not know the level of diffusivity in the disc (this will be a topic of interest in Section~\ref{sec:results}), nor the exact value of $\Sigma_{\rm gas}$, we infer the scaling factor for large dust, $\chi_{\rm ld}$, from the width of Ring13. In the radial profile this ring is observed two to three times wider in the gas-tracing NIR than in the fluxes received from larger grains by ALMA. We assume that the same ratio holds in the vertical direction (due to the settling of larger grains towards the mid-plane of the disc), leading to $\chi_{\mathrm{ld}}=0.4$. This is analogous to assuming equal radial and vertical turbulent diffusions.

For the vertical structure, \citet{dOrazi} found flaring angles (i.e. height of the ring over the disc midplane divided by the radius of the ring), of $\varphi$ = 6.2$\pm$0.6\,deg for the inner ring and $\varphi$ = 8.5$\pm$1.0\,deg for the outer one. Our model aims to reproduce those values by using two different flaring indices, $\psi_1$ and $\psi_2$. The separation between the two values was set at $r = 18$\,au with $\psi_1=0.2$ for Ring5 and Ring13, and $\psi_2=0.5$ for Ring24. The scale height is set to $H_\circ = 0.89$\,au at $r_\circ = 18$\,au. This vertical structure is consistent with the measurements made by \citet{dOrazi}, as the model flaring angles are $\varphi_{\mathrm{inner}}$ = 6.4\,deg and $\varphi_{\mathrm{outer}}$ = 7.6\,deg. This is summarized in the bottom panel of Fig.~\ref{fig:profiles} that shows the aspect ratio profile $h(r)=H(r)/r$.

Although both ALMA and SPHERE--IRDIS images display two-ringed morphologies, we propose a three-ringed structure plus an inner disc to reproduce the observations. This bold decision is due the major fit improvement in the SED and the polarize image (more on this in Sec.~\ref{sec:results}). We separate the disc into four individual regions: the inner disc, Ring5, Ring13, and Ring24. The combined gas surface density profile is then given by:
\begin{equation}
  \Sigma_{\mathrm{gas}}(r) = \Sigma_{\mathrm{inner\,disc}}(r) + \Sigma_{\mathrm{R5}}(r) + \Sigma_{\mathrm{R13}}(r) + \Sigma_{\mathrm{R24}}(r)
\end{equation}

@@

We chose to use Gaussian profiles to parametrize Ring5 and Ring13,
\begin{equation}
  \Sigma_{\mathrm{ring}}(r) = \frac{\Sigma_\circ}{\sqrt{2 \pi} \sigma}
  \, \mathrm{exp}\left[-\frac{1}{2}\left(\frac{r-\mu}{\sigma}\right)^{2}\right],
\end{equation}
where we define constants that correspond to the centroid radii \{$\mu_{\mathrm{R5}}=5.2$\,au, $\mu_{\mathrm{R13}}=14.9$\,au\}, ring widths \{$\sigma_{\mathrm{R5}}=0.25$\,au, $\sigma_{\mathrm{R13}}=2.26$\,au\}, and normalizations \{$\Sigma_{\circ,\mathrm{R5}}=3.3$\,$\mathrm{g\,cm^{-3}}$, $\Sigma_{\circ,\mathrm{R13}}=6.0 \times 10^{-1}$\,$\mathrm{g\,cm^{-3}}$\} for both components separately.

For Ring24 we used the same power-law as for the inner disc but scaled by an empirically obtained factor, $\delta_{\mathrm{sd}}(r)$ and by $\epsilon(r)$, a parameter that allows us model a smoother inner edge of the outer ring:
\begin{equation}
  \Sigma_{\mathrm{R24}}(r) = \Sigma_{\mathrm{inner\,disc}}(r)\, \delta_{\mathrm{sd}}(r)\,\epsilon(r),
\end{equation}
with $\delta_{\mathrm{sd}}(r) = 1.0\times 10^5$ for $r > 18$\,au and zero for lower radii, and
%\begin{equation}
%  \delta_{\mathrm{sd}}(r) =
%  \begin{cases}
%  0    & r < 18\text{\,au} \\
%  4.2\times 10^4 & r > 18\text{\,au},
%  \end{cases}
%\end{equation}
\begin{equation}
    \epsilon(r) = 
    \begin{cases}
  1    & r < R_\mathrm{in} \,\, {\rm and} \,\, R_\mathrm{peak} < r \\
  \left(\frac{ r - R_\mathrm{in}}{R_\mathrm{peak} - R_\mathrm{in}}\right)^3 & R_\mathrm{in} < r < R_\mathrm{peak},
    \end{cases}
\end{equation}
where $R_\mathrm{in}$ and $R_\mathrm{peak}$ respectively mark the inner edge and the location of maximum density of the outer ring. We used $R_\mathrm{in,gas}$ = 18.0\,au and $R_\mathrm{peak,gas}$ = 26.4\,au.

The total dust-to-gas mass ratio is taken to be $\zeta = 0.047$ \citep[as in][]{Rosenfeld_2013}. The small dust grains are assumed to only make up for a fraction of $f_\mathrm{sd}=1\%$ of the total dust mass. As small dust is typically tightly coupled to the gas dynamics, its density profile is expected to follow the gas density. Then the density of small dust can be calculated as:
\begin{equation}
\rho_{\mathrm{small dust}}(r,z)=\rho_{\mathrm{gas}}(r,z)\, f_{\mathrm{sd}} \: \zeta .
\end{equation}

Since the large dust grains are less coupled to the gas, their distribution has some important differences that require a special parameterisation, with a larger inner cavity, a larger gap between Ring13 and Ring24, and a break in the outer ring. We only included a low density of large grains within Ring5, just underneath the detection limit of the ALMA observation, as it does not show any visible signature. We also truncated the profile of large dust grains at 63\,au. It is then defined by the sum of its three components
\begin{equation}
  \Sigma_{\mathrm{ld}}(r) = \Sigma_{\mathrm{R5,ld}} + 
  \Sigma_{\mathrm{R13,ld}} + \Sigma_{\mathrm{R24,ld}}.
\end{equation}
